My Title

### 2025-12-09 misc/JSS article/final/article.Rnw

While this article cannot fully present \pkg{collapse}, the following sections introduce its key features, starting with (\ref{sec:fast_stat_fun}) the \emph{Fast Statistical Functions} and their (\ref{sec:integration}) integration with data manipulation functions; (\ref{sec:ts_ps}) architecture for time series and panel data; (\ref{sec:join_pivot}) table joins and pivots; (\ref{sec:list_proc}) list processing functions; (\ref{sec:summ_stat}) descriptive tools; and (\ref{sec:glob_opt}) global options. Section~\ref{sec:bench} provides a small benchmark, Section~\ref{sec:conclusion} concludes. For deeper engagement with \pkg{collapse}, consult the \href{https://fastverse.github.io/collapse/articles/collapse_documentation.html}{documentation resources}, including \href{https://fastverse.github.io/collapse/articles/index.html}{vignettes}, \href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/collapse\%20cheat\%20sheet/collapse_cheat_sheet.pdf}{cheatsheet}, \href{https://sebkrantz.github.io/Rblog/}{blog}, and \href{https://www.youtube.com/watch?v=OwWT1-dSEts}{talk}$+$\href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{slides}.

@@

Grouped programming using 'GRP' objects and \emph{Fast Statistical Functions} is also particularly powerful with vectors and matrices. For example, in the \href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{useR 2022 presentation}, I aggregate 32 global input-output tables stored as matrices (\code{x}) from the country to the region level using a single grouping object (\code{g}) and expressions like \code{x |> fsum(g) |> t() |> fsum(g) |> t()}---computing 45 million sums crunching 5.7GB of data in $\sim$0.3 seconds on my M1.\footnote{Another recent application with vectors involved numerically optimizing a parameter $a$ in an equation of the form $\sum_i x_{ij}^a\ \forall j\in J$ so as to minimize the deviation from a target $y_j$ where there are $J$ groups (1 million in my case)---see the first example in \href{https://sebkrantz.github.io/Rblog/2023/04/12/collapse-and-the-fastverse-reflecting-the-past-present-and-future/}{this blog post} for an illustration.}

@@

An example follows using the \code{exports} data. \emph{Note} that data can be unsorted for indexation.
<<indexing>>=
exportsi <- exports |> findex_by(c, s, y)
exportsi |> G(0:1) |> head(5)
exportsi |> findex() |> print(2)
@
The index statistics are: \code{[N. ids] | [N. periods (total periods: (max-min)/GCD)]}. % This extracts an \class{indexes\_series} and demonstrates its properties.
<<indexing_2>>=
vi <- exportsi$v; str(vi, width = 70, strict = "cut")
is_irregular(vi)
vi |> psmat() |> head(3)
fdiff(vi) |> psmat() |> head(3)
@
\fct{psmat}, for panel-series to matrix, generates a matrix or array from panel data. Thanks to deep indexation, indexed computations work inside arbitrary data masking environments:
<<indexing_3>>=
settransform(exportsi, v_ld = Dlog(v))
lm(v_ld ~ L(v_ld, 1:2), exportsi) |> summary() |> coef() |> round(3)
@
Indexed series/frames also support transformations such as grouped scaling with \fct{fscale} or demeaning with \fct{fwithin}. Functions \fct{psacf}/\fct{pspacf}/\fct{psccf} provide panel-data autocorrelation functions, which are computed using group-scaled and suitably lagged panel-series. The \class{index\_df} attached to these objects can also be used with other general tools such as \fct{collapse::BY} to perform grouped computations using third-party functions: %For example, below I compute a 5-year rolling average ().
%
<<rollmean>>=
BY(vi, findex(vi)$c.s, data.table::frollmean, 5) |> head(10)
@
%
Last but not least, the computational performance of these classes is second to none.\footnote{See, e.g., the small benchmark presented on slide 40 of the \href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{useR 2022 presentation}.}
%
\section{Table joins and pivots} \label{sec:join_pivot}
%
Among its suite of \href{https://fastverse.github.io/collapse/reference/fast-data-manipulation.html}{data manipulation functions}, \pkg{collapse}'s implementations of table \href{https://fastverse.github.io/collapse/reference/join.html}{joins} and \href{https://fastverse.github.io/collapse/reference/pivot.html}{pivots} are particularly noteworthy since they offer several new features, including rich verbosity for table joins, pivots supporting variable labels, and recast pivots. Both implementations also offer outstanding computational performance, syntax, and memory efficiency.
%
\subsection{Joins}
%
Compared to commercial software such as \proglang{STATA} \citep{STATA}, the implementation of joins in most open-source software provides no information on how many records were joined from both tables. This often provokes manual efforts to validate the join operation. \fct{collapse::join} provides many options to understand table join operations. Its syntax is:
\begin{Code}
join(x, y, on = NULL, how = "left", suffix = NULL, validate = "m:m",
  multiple = FALSE, sort = FALSE, keep.col.order = TRUE, verbose = 1,
  drop.dup.cols = FALSE, require = NULL, column = NULL, attr = NULL, ...)
\end{Code}
It defaults to left join and only takes first matches from \code{y} (\code{multiple = FALSE}). Thus, it simply adds columns to \code{x}, which is efficient and sufficient/desired in many cases. By default (\code{verbose = 1}), it prints information about the join operation and number of records joined. \newline

>>>>

While this article cannot fully present \pkg{collapse}, the following sections introduce its key features, starting with (\ref{sec:fast_stat_fun}) the \emph{Fast Statistical Functions} and their (\ref{sec:integration}) integration with data manipulation functions; (\ref{sec:ts_ps}) architecture for time series and panel data; (\ref{sec:join_pivot}) table joins and pivots; (\ref{sec:list_proc}) list processing functions; (\ref{sec:summ_stat}) descriptive tools; and (\ref{sec:glob_opt}) global options. Section~\ref{sec:bench} provides a small benchmark, Section~\ref{sec:conclusion} concludes. For deeper engagement with \pkg{collapse}, consult the \href{https://fastverse.github.io/collapse/articles/collapse_documentation.html}{documentation resources}, including \href{https://fastverse.github.io/collapse/articles/index.html}{vignettes}, \href{https://raw.githubusercontent.com/fastverse/collapse/master/misc/collapse\%20cheat\%20sheet/collapse_cheat_sheet.pdf}{cheatsheet}, \href{https://sebkrantz.github.io/Rblog/}{blog}, and \href{https://www.youtube.com/watch?v=OwWT1-dSEts}{talk}$+$\href{https://raw.githubusercontent.com/fastverse/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{slides}.

@@

Grouped programming using 'GRP' objects and \emph{Fast Statistical Functions} is also particularly powerful with vectors and matrices. For example, in the \href{https://raw.githubusercontent.com/fastverse/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{useR 2022 presentation}, I aggregate 32 global input-output tables stored as matrices (\code{x}) from the country to the region level using a single grouping object (\code{g}) and expressions like \code{x |> fsum(g) |> t() |> fsum(g) |> t()}---computing 45 million sums crunching 5.7GB of data in $\sim$0.3 seconds on my M1.\footnote{Another recent application with vectors involved numerically optimizing a parameter $a$ in an equation of the form $\sum_i x_{ij}^a\ \forall j\in J$ so as to minimize the deviation from a target $y_j$ where there are $J$ groups (1 million in my case)---see the first example in \href{https://sebkrantz.github.io/Rblog/2023/04/12/collapse-and-the-fastverse-reflecting-the-past-present-and-future/}{this blog post} for an illustration.}

@@

An example follows using the \code{exports} data. \emph{Note} that data can be unsorted for indexation.
<<indexing>>=
exportsi <- exports |> findex_by(c, s, y)
exportsi |> G(0:1) |> head(5)
exportsi |> findex() |> print(2)
@
The index statistics are: \code{[N. ids] | [N. periods (total periods: (max-min)/GCD)]}. % This extracts an \class{indexes\_series} and demonstrates its properties.
<<indexing_2>>=
vi <- exportsi$v; str(vi, width = 70, strict = "cut")
is_irregular(vi)
vi |> psmat() |> head(3)
fdiff(vi) |> psmat() |> head(3)
@
\fct{psmat}, for panel-series to matrix, generates a matrix or array from panel data. Thanks to deep indexation, indexed computations work inside arbitrary data masking environments:
<<indexing_3>>=
settransform(exportsi, v_ld = Dlog(v))
lm(v_ld ~ L(v_ld, 1:2), exportsi) |> summary() |> coef() |> round(3)
@
Indexed series/frames also support transformations such as grouped scaling with \fct{fscale} or demeaning with \fct{fwithin}. Functions \fct{psacf}/\fct{pspacf}/\fct{psccf} provide panel-data autocorrelation functions, which are computed using group-scaled and suitably lagged panel-series. The \class{index\_df} attached to these objects can also be used with other general tools such as \fct{collapse::BY} to perform grouped computations using third-party functions: %For example, below I compute a 5-year rolling average ().
%
<<rollmean>>=
BY(vi, findex(vi)$c.s, data.table::frollmean, 5) |> head(10)
@
%
Last but not least, the computational performance of these classes is second to none.\footnote{See, e.g., the small benchmark presented on slide 40 of the \href{https://raw.githubusercontent.com/fastverse/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{useR 2022 presentation}.}
%
\section{Table joins and pivots} \label{sec:join_pivot}
%
Among its suite of \href{https://fastverse.github.io/collapse/reference/fast-data-manipulation.html}{data manipulation functions}, \pkg{collapse}'s implementations of table \href{https://fastverse.github.io/collapse/reference/join.html}{joins} and \href{https://fastverse.github.io/collapse/reference/pivot.html}{pivots} are particularly noteworthy since they offer several new features, including rich verbosity for table joins, pivots supporting variable labels, and recast pivots. Both implementations also offer outstanding computational performance, syntax, and memory efficiency.
%
\subsection{Joins}
%
Compared to commercial software such as \proglang{STATA} \citep{STATA}, the implementation of joins in most open-source software provides no information on how many records were joined from both tables. This often provokes manual efforts to validate the join operation. \fct{collapse::join} provides many options to understand table join operations. Its syntax is:
\begin{Code}
join(x, y, on = NULL, how = "left", suffix = NULL, validate = "m:m",
  multiple = FALSE, sort = FALSE, keep.col.order = TRUE, verbose = 1,
  drop.dup.cols = FALSE, require = NULL, column = NULL, attr = NULL, ...)
\end{Code}
It defaults to left join and only takes first matches from \code{y} (\code{multiple = FALSE}). Thus, it simply adds columns to \code{x}, which is efficient and sufficient/desired in many cases. By default (\code{verbose = 1}), it prints information about the join operation and number of records joined. \newline

### 2025-12-09 misc/JSS article/final/article.tex

Indexed by:  c.s [2] | y [9 (10)]
\end{Soutput}
\end{Schunk}
%
Last but not least, the computational performance of these classes is second to none.\footnote{See, e.g., the small benchmark presented on slide 40 of the \href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{useR 2022 presentation}.}
%
\section{Table joins and pivots} \label{sec:join_pivot}
%
Among its suite of \href{https://fastverse.github.io/collapse/reference/fast-data-manipulation.html}{data manipulation functions}, \pkg{collapse}'s implementations of table \href{https://fastverse.github.io/collapse/reference/join.html}{joins} and \href{https://fastverse.github.io/collapse/reference/pivot.html}{pivots} are particularly noteworthy since they offer several new features, including rich verbosity for table joins, pivots supporting variable labels, and recast pivots. Both implementations also offer outstanding computational performance, syntax, and memory efficiency.
%
\subsection{Joins}
%
Compared to commercial software such as \proglang{STATA} \citep{STATA}, the implementation of joins in most open-source software provides no information on how many records were joined from both tables. This often provokes manual efforts to validate the join operation. \fct{collapse::join} provides many options to understand table join operations. Its syntax is:
\begin{Code}
join(x, y, on = NULL, how = "left", suffix = NULL, validate = "m:m",
  multiple = FALSE, sort = FALSE, keep.col.order = TRUE, verbose = 1,
  drop.dup.cols = FALSE, require = NULL, column = NULL, attr = NULL, ...)
\end{Code}
It defaults to left join and only takes first matches from \code{y} (\code{multiple = FALSE}). Thus, it simply adds columns to \code{x}, which is efficient and sufficient/desired in many cases. By default (\code{verbose = 1}), it prints information about the join operation and number of records joined. \newline

>>>>

Indexed by:  c.s [2] | y [9 (10)]
\end{Soutput}
\end{Schunk}
%
Last but not least, the computational performance of these classes is second to none.\footnote{See, e.g., the small benchmark presented on slide 40 of the \href{https://raw.githubusercontent.com/fastverse/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{useR 2022 presentation}.}
%
\section{Table joins and pivots} \label{sec:join_pivot}
%
Among its suite of \href{https://fastverse.github.io/collapse/reference/fast-data-manipulation.html}{data manipulation functions}, \pkg{collapse}'s implementations of table \href{https://fastverse.github.io/collapse/reference/join.html}{joins} and \href{https://fastverse.github.io/collapse/reference/pivot.html}{pivots} are particularly noteworthy since they offer several new features, including rich verbosity for table joins, pivots supporting variable labels, and recast pivots. Both implementations also offer outstanding computational performance, syntax, and memory efficiency.
%
\subsection{Joins}
%
Compared to commercial software such as \proglang{STATA} \citep{STATA}, the implementation of joins in most open-source software provides no information on how many records were joined from both tables. This often provokes manual efforts to validate the join operation. \fct{collapse::join} provides many options to understand table join operations. Its syntax is:
\begin{Code}
join(x, y, on = NULL, how = "left", suffix = NULL, validate = "m:m",
  multiple = FALSE, sort = FALSE, keep.col.order = TRUE, verbose = 1,
  drop.dup.cols = FALSE, require = NULL, column = NULL, attr = NULL, ...)
\end{Code}
It defaults to left join and only takes first matches from \code{y} (\code{multiple = FALSE}). Thus, it simply adds columns to \code{x}, which is efficient and sufficient/desired in many cases. By default (\code{verbose = 1}), it prints information about the join operation and number of records joined. \newline

### 2025-12-09 misc/JSS article/final/jss5278/jss5278.tex

This article cannot fully present \pkg{collapse}, but the following sections introduce its key features, starting with (\ref{sec:fast_stat_fun}) the \emph{Fast Statistical Functions} and their (\ref{sec:integration}) integration with data manipulation functions; (\ref{sec:ts_ps}) architecture for time series and panel data; (\ref{sec:join_pivot}) table joins and pivots; (\ref{sec:list_proc}) list processing functions; (\ref{sec:summ_stat}) descriptive tools; and (\ref{sec:glob_opt}) global options. Section~\ref{sec:bench} provides a small benchmark, Section~\ref{sec:conclusion} concludes. For deeper engagement with \pkg{collapse}, consult the \href{https://fastverse.github.io/collapse/articles/collapse_documentation.html}{documentation resources}, including \href{https://fastverse.github.io/collapse/articles/index.html}{vignettes}, \href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/collapse\%20cheat\%20sheet/collapse_cheat_sheet.pdf}{cheatsheet}, \href{https://sebkrantz.github.io/Rblog/}{blog}, and \href{https://www.youtube.com/watch?v=OwWT1-dSEts}{talk} with \href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{slides}.

>>>>

This article cannot fully present \pkg{collapse}, but the following sections introduce its key features, starting with (\ref{sec:fast_stat_fun}) the \emph{Fast Statistical Functions} and their (\ref{sec:integration}) integration with data manipulation functions; (\ref{sec:ts_ps}) architecture for time series and panel data; (\ref{sec:join_pivot}) table joins and pivots; (\ref{sec:list_proc}) list processing functions; (\ref{sec:summ_stat}) descriptive tools; and (\ref{sec:glob_opt}) global options. Section~\ref{sec:bench} provides a small benchmark, Section~\ref{sec:conclusion} concludes. For deeper engagement with \pkg{collapse}, consult the \href{https://fastverse.github.io/collapse/articles/collapse_documentation.html}{documentation resources}, including \href{https://fastverse.github.io/collapse/articles/index.html}{vignettes}, \href{https://raw.githubusercontent.com/fastverse/collapse/master/misc/collapse\%20cheat\%20sheet/collapse_cheat_sheet.pdf}{cheatsheet}, \href{https://sebkrantz.github.io/Rblog/}{blog}, and \href{https://www.youtube.com/watch?v=OwWT1-dSEts}{talk} with \href{https://raw.githubusercontent.com/fastverse/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{slides}.

### 2025-12-09 misc/collapse cheat sheet/collapse_cheat_sheet.Rnw

% \end{adjustbox}
%}
% \hrrule
\vspace{-5mm}
\textcolor{lightgray}{\hrulefill}\\
{\scriptsize \vspace{-0.5mm}
 Page 1 of 2 \hfill \href{https://creativecommons.org/licenses/by-sa/4.0/}{CC-BY-SA}\ Sebastian Krantz\ \textbullet\ Learn more at \href{https://fastverse.github.io/collapse/}{fastverse.github.io/collapse}\ \textbullet\ Source code at \href{https://github.com/SebKrantz/collapse}{github.com/SebKrantz/collapse}\ \textbullet\ Updates announced at \href{https://twitter.com/collapse\_R}{twitter.com/collapse\_R} - \#rcollapse\ \textbullet\ Cheatsheet created for \emph{collapse} version 2.0.3\ \textbullet\ Updated: 2023-10
}

@@

\vspace{-5.5mm}
\textcolor{lightgray}{\hrulefill}\\
{\scriptsize \vspace{-0.5mm}
 Page 2 of 2 \hfill \href{https://creativecommons.org/licenses/by-sa/4.0/}{CC-BY-SA}\ Sebastian Krantz\ \textbullet\ Learn more at \href{https://fastverse.github.io/collapse/}{fastverse.github.io/collapse}\ \textbullet\ Source code at \href{https://github.com/SebKrantz/collapse}{github.com/SebKrantz/collapse}\ \textbullet\ Updates announced at \href{https://twitter.com/collapse\_R}{twitter.com/collapse\_R} - \#rcollapse\ \textbullet\ Cheatsheet created for \emph{collapse} version 2.0.3\ \textbullet\ Updated: 2023-10
}

>>>>

% \end{adjustbox}
%}
% \hrrule
\vspace{-5mm}
\textcolor{lightgray}{\hrulefill}\\
{\scriptsize \vspace{-0.5mm}
 Page 1 of 2 \hfill \href{https://creativecommons.org/licenses/by-sa/4.0/}{CC-BY-SA}\ Sebastian Krantz\ \textbullet\ Learn more at \href{https://fastverse.github.io/collapse/}{fastverse.github.io/collapse}\ \textbullet\ Source code at \href{https://github.com/fastverse/collapse}{github.com/fastverse/collapse}\ \textbullet\ Updates announced at \href{https://twitter.com/collapse\_R}{twitter.com/collapse\_R} - \#rcollapse\ \textbullet\ Cheatsheet created for \emph{collapse} version 2.0.3\ \textbullet\ Updated: 2023-10
}

@@

\vspace{-5.5mm}
\textcolor{lightgray}{\hrulefill}\\
{\scriptsize \vspace{-0.5mm}
 Page 2 of 2 \hfill \href{https://creativecommons.org/licenses/by-sa/4.0/}{CC-BY-SA}\ Sebastian Krantz\ \textbullet\ Learn more at \href{https://fastverse.github.io/collapse/}{fastverse.github.io/collapse}\ \textbullet\ Source code at \href{https://github.com/fastverse/collapse}{github.com/fastverse/collapse}\ \textbullet\ Updates announced at \href{https://twitter.com/collapse\_R}{twitter.com/collapse\_R} - \#rcollapse\ \textbullet\ Cheatsheet created for \emph{collapse} version 2.0.3\ \textbullet\ Updated: 2023-10
}
